Substructure Boosts to Dark Matter Annihilation from Sommerfeld Enhancement

### 2009-02-27 tex/somm.tex
Updated figures + text

\newcommand{\eg}{e.g.}
\newcommand{\ie}{i.e.}
\newcommand{\Fermi}{\emph{Fermi}}
\newcommand{\VL}{Via Lactea}
\newcommand{\aquarius}{Aquarius}
\newcommand{\CDM}{CDM}
\newcommand{\LCDM}{\ensuremath{\Lambda}\CDM}
\newcommand{\NFW}{NFW}
\newcommand{\GNFW}{G\NFW}
\newcommand{\DM}{DM}
\newcommand{\SM}{SM}
\newcommand{\OM}{OM}
\newcommand{\mpl}{\ensuremath{m_{\mathrm{Pl}}}}
\newcommand{\somm}{\ensuremath{S}}
\newcommand{\mdm}{\ensuremath{m_{\chi}}}
\newcommand{\mv}{\ensuremath{m_{\phi}}}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\Eqnname}{Equation}
\newcommand{\eqnname}{equation}
\newcommand{\Ngamma}{\ensuremath{N_{\gamma}}}
\newcommand{\Ngammai}{\ensuremath{N_{\gamma,i}}}
\newcommand{\Eth}{\ensuremath{E_{\mathrm{th}}}}
\newcommand{\sigmaannv}{\ensuremath{\langle\sigma v\rangle}}
\newcommand{\los}{los}
\newcommand{\order}{\ensuremath{\mathcal{O}}}
\newcommand{\sigv}{\ensuremath{\sigma_v}}
\newcommand{\lum}{\ensuremath{\mathcal{L}}}
\newcommand{\lumsmooth}{\ensuremath{\lum_{\mbox{{\footnotesize sm}}}}}
\newcommand{\lumsmoothc}{\ensuremath{\lum_{\mbox{{\footnotesize sm}},v=c}}}
\newcommand{\lumsmoothtoo}{\ensuremath{\overline{\mathcal{L}}}}
\newcommand{\lumsub}{\ensuremath{\lum_{\mbox{{\footnotesize sub}}}}}
\newcommand{\boost}{\ensuremath{B}}
\newcommand{\redboost}{\ensuremath{\mathcal{B}_{\mbox{{\footnotesize red}}}}}
\newcommand{\totalboost}{\ensuremath{\mathcal{B}}}
\newcommand{\redr}{\ensuremath{\tilde{r}}}
\newcommand{\rhos}{\ensuremath{\rho_s}}
\newcommand{\rs}{\ensuremath{r_s}}
\newcommand{\dist}{\ensuremath{D}}
\newcommand{\norm}{\ensuremath{\mathcal{N}}}
\newcommand{\normnfw}{\ensuremath{\norm_1}}
\newcommand{\normeinasto}{\ensuremath{\norm_2}}
\newcommand{\rhominustwo}{\ensuremath{\rho_{-2}}}
\newcommand{\rminustwo}{\ensuremath{r_{-2}}}
\newcommand{\sommboost}{\ensuremath{\mathcal{S}}}
\newcommand{\clumpboost}{\ensuremath{\mathcal{C}}}
\newcommand{\alphaEinasto}{\ensuremath{\alpha}}
\newcommand{\Msol}{\ensuremath{M_{\odot}}}
\newcommand{\Msun}{\Msol}
\newcommand{\Mvir}{\ensuremath{M_{\mbox{{\footnotesize vir}}}}}
\newcommand{\Rvir}{\ensuremath{r_{\mbox{{\footnotesize vir}}}}}
\newcommand{\conc}{\ensuremath{c}}
\newcommand{\rhoc}{\ensuremath{\rho_{\mbox{{\footnotesize c}}}}}
\newcommand{\omegam}{\ensuremath{\Omega_{m}}}
\newcommand{\vcirc}{\ensuremath{v_{\mbox{{\footnotesize c}}}}}
\newcommand{\vmax}{\ensuremath{v_{\mbox{{\footnotesize max}}}}}
\newcommand{\rmax}{\ensuremath{r_{\mbox{{\footnotesize max}}}}}
\newcommand{\dSph}{dSph}
\newcommand{\dSphs}{dSphs}
\newcommand{\SDSS}{SDSS}
\newcommand{\degree}{^\circ}

@@

In this paper we quantify the magnitude of this new substructure boost
and discuss its implications for the detection of dark matter
annihilation from subhalos in the Galactic halos. As we will show, the
total combined boost from Sommerfeld enhancement and substructure can
reach very large values (10$^5$ -- 10$^9$), and these boosts are only
weakly dependent on the assumptions that we have to make about the
abundance of substructure and the \DM\ density profile. We use these
total boosts together with published \DM\ density profiles of some of
the dwarf Spheroidal satellite galaxies of the Milky Way to predict
the total flux of gamma-rays from \DM\ annihilation that would be
detectable by Fermi and find that in most of the models that posit the
Sommerfeld enhancement the \DM\ annihilation signal from the dwarf
Spheroidals shines above the diffuse extragalactic gamma-ray
background, which is the only background signal at the high Galactic
lattitudes of these satellite galaxies. Finally, we also consider the
prospects for detecting smaller subhalos of the Galaxy by computing
annihilation fluxes of subhalos in the \VL\-2 simulation. We will see
that whether these subhalos will be detectable in the near future is
highly dependent on the exact details of the Sommerfeld enhancement.

@@

\begingroup
%\squeezetable
\begin{table*}
\begin{ruledtabular}
\begin{tabular}{lrr@{.}lr@{}lr@{}lr@{.}lr@{.}lr@{.}l@{ $\times$ }lc}
dSph & \dist\ & \multicolumn{2}{c}{\sigv}\ & \multicolumn{2}{c}{\rmax} & \multicolumn{2}{c}{\vmax} & \multicolumn{2}{c}{\rhos} & \multicolumn{2}{c}{\rs} & \multicolumn{3}{c}{$M(<\rs)$} & References\\
& [kpc] & \multicolumn{2}{c}{[km s$^{-1}$]} & \multicolumn{2}{c}{[kpc]} & \multicolumn{2}{c}{[km s$^{-1}$]} & \multicolumn{2}{c}{[$\Msun$ kpc$^{-3}$]} & \multicolumn{2}{c}{[kpc]} & \multicolumn{3}{c}{[$\Msun$]}&\\[4pt]
\hline
Carina.......................................................................... & 101 & 6&8 & 3&\phantom{f}      & 16&$\!\!\!$      & 1&1 $\times$ $10^7$ & 1&4 & 7&4&$10^8$ & 1,3\\
Coma Berenices...........................................................      & 44  & 4&6 & $\cdot$&& $\cdot$& & 3&0 $\times$ $10^8$ & 0&3 & 2&0&10$^7$ & 2,4\\
Draco........................................................................... & 80  & 5&5 & 8&      & 40&      & 1&0 $\times$ $10^7$ & 3&7 & 1&2&$10^9$ & 1,3\\
Fornax.........................................................................  & 138 & 11&1& 4&      & 20&      & 1&0 $\times$ $10^7$ & 1&9 & 1&5&$10^8$ & 1,3\\
Leo I............................................................................  & 250 & 8&8 & 6&      & 30&      & 1&0 $\times$ $10^7$ & 2&8 & 5&2&$10^8$ & 1,3\\
Leo II...........................................................................  & 205 & 6&7 & 4&      & 20&      & 1&0 $\times$ $10^7$ & 1&9 & 1&5&$10^8$ & 1,3\\
Sculptor.......................................................................  & 79  & 8&5 & 6&      & 30&      & 1&0 $\times$ $10^7$ & 2&8 & 5&2&$10^8$ & 1,3\\
Sextans........................................................................  & 86  & 5&8 & 2&      & 10&      & 1&0 $\times$ $10^7$ & 0&9 & 1&9&$10^7$ & 1,3\\
Ursa Major II..............................................................   & 32  & 6&7 & $\cdot$&& $\cdot$& & 3&0 $\times$ $10^8$ & 0&3 & 2&0&$10^7$ & 2,4\\
Ursa Minor..................................................................    & 66  & 15&0& 6&      & 30&      & 1&0 $\times$ $10^8$ & 2&8 & 5&2&$10^8$ & 1,3\\
%Willman I............. & 38 &&& 5&5 $\times$ $10^8$ & 0&2 & 1&10$^{15}$\\
%Segue I.................. & 23 &&&&
\end{tabular}
\end{ruledtabular}
\caption{Properties of the dwarf Spheroidals used in this study.\\
References: (1) \citet{2008ApJ...672..904P}, (2) \citet{2008ApJ...678..614S}, (3) \citet{1998ARA&A..36..435M}, (4) \citet{2007ApJ...670..313S}.}\label{tab:dsphs}
\end{table*}
\endgroup

@@

It is clear from Figure \ref{fig:dsph} that the \DM\ annihilation flux
predicted to be coming from the \dSphs\ is very large in a large class
of models. We have to stress here that our predicted fluxes are all
for the particle physics specific parameters given in Section
\ref{sec:partphys}, \ie, \sigmaannv\ = $3 \times 10^{-26}$ cm$^3$
s$^{-1}$, \mdm\ = 700 GeV and a particular form for the \DM\
annihilation spectrum. To predict the gamma-ray flux for a particular
model with different values of these particle physics parameters one
is required to calculate the particle physics parameter specific
factor to the \DM\ annihilation flux and appropriately rescale the
values given here. In order to compare these signals to the expected
background flux of extragalactic gamma rays (which is the only
relevant background since these \dSphs\ are at high Galactic
lattitude), we have integrated the extragalactic gamma ray flux as
measured by EGRET over the energy range between 4 GeV -- 250 GeV and
multiplied it with the solid angle that the \dSphs\ span on the
sky. The extragalactic gamma ray flux is given by a power law
\cite{1998ApJ...494..523S}
\begin{equation}
\frac{\dd \Ngamma}{\dd E \,\dd A \, \dd t\,\dd \Omega} = k\,\left(\frac{E}{E_0}\right)^{-\alpha}\, ,
\end{equation}
where $k = 7.3 \times 10^{-9}$, $\alpha$ = 2.10, and $E_0 = 0.45$
GeV. We have included the background flux of extragalactic photons in
Figure \ref{fig:dsph}. This background flux is in all cases of the
order of 10$^{-13}$ photons cm$^{-2}$ s$^{-1}$, such that the \DM\
annihilation boosted by substructure and Sommerfeld enhancement
outshine this background in all of the considered models, and by many
orders of magnitude in most models. Given that the flux above 100 MeV
over a 55 days time-span needed to separate a point source of \DM\
annihilation from the background by \Fermi\ to give a 5$\sigma$
detection is about 10$^{-8}$ photons cm$^{-1}$ s$^{-1}$ and that the
necessary fluxes scale roughly as $t_{\mbox{{\footnotesize
exp}}}^{-1/2}$ \cite{2008JCAP...07..013B}, the \DM\ annihilation of
these \dSphs\ should be detectable by \Fermi\ in many of the models
parameterized by ($\mv/\mdm,\alpha$). We conclude that the prospects
for detecting \DM\ annihilation from the \dSphs\ with \Fermi\ are very
good in models that include a significant Sommerfeld enhancement.

@@

{\bf Acknowledgements:} It is a pleasure to thank Mulin Ding, David
Hogg, and Neal Weiner for comments and assistance. This research was partially
supported by NASA (grant NNX08AJ48G).

>>>>

\newcommand{\eg}{e.g.}
\newcommand{\ie}{i.e.}
\newcommand{\Fermi}{\emph{Fermi}}
\newcommand{\VL}{Via Lactea}
\newcommand{\aquarius}{Aquarius}
\newcommand{\CDM}{CDM}
\newcommand{\LCDM}{\ensuremath{\Lambda}\CDM}
\newcommand{\NFW}{NFW}
\newcommand{\GNFW}{G\NFW}
\newcommand{\DM}{DM}
\newcommand{\SM}{SM}
\newcommand{\OM}{OM}
\newcommand{\mpl}{\ensuremath{m_{\mathrm{Pl}}}}
\newcommand{\somm}{\ensuremath{S}}
\newcommand{\mdm}{\ensuremath{m_{\chi}}}
\newcommand{\mv}{\ensuremath{m_{\phi}}}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\Eqnname}{Equation}
\newcommand{\eqnname}{equation}
\newcommand{\Ngamma}{\ensuremath{N_{\gamma}}}
\newcommand{\Ngammai}{\ensuremath{N_{\gamma,i}}}
\newcommand{\Eth}{\ensuremath{E_{\mathrm{th}}}}
\newcommand{\sigmaannv}{\ensuremath{\langle\sigma v\rangle}}
\newcommand{\los}{los}
\newcommand{\order}{\ensuremath{\mathcal{O}}}
\newcommand{\sigv}{\ensuremath{\sigma_v}}
\newcommand{\lum}{\ensuremath{\mathcal{L}}}
\newcommand{\lumsmooth}{\ensuremath{\lum_{\mbox{{\footnotesize sm}}}}}
\newcommand{\lumsmoothc}{\ensuremath{\lum_{\mbox{{\footnotesize sm}},v=c}}}
\newcommand{\lumsmoothtoo}{\ensuremath{\overline{\mathcal{L}}}}
\newcommand{\lumsub}{\ensuremath{\lum_{\mbox{{\footnotesize sub}}}}}
\newcommand{\boost}{\ensuremath{B}}
\newcommand{\redboost}{\ensuremath{\mathcal{B}_{\mbox{{\footnotesize red}}}}}
\newcommand{\totalboost}{\ensuremath{\mathcal{B}}}
\newcommand{\redr}{\ensuremath{\tilde{r}}}
\newcommand{\rhos}{\ensuremath{\rho_s}}
\newcommand{\rs}{\ensuremath{r_s}}
\newcommand{\dist}{\ensuremath{D}}
\newcommand{\norm}{\ensuremath{\mathcal{N}}}
\newcommand{\normnfw}{\ensuremath{\norm_1}}
\newcommand{\normeinasto}{\ensuremath{\norm_2}}
\newcommand{\rhominustwo}{\ensuremath{\rho_{-2}}}
\newcommand{\rminustwo}{\ensuremath{r_{-2}}}
\newcommand{\sommboost}{\ensuremath{\mathcal{S}}}
\newcommand{\clumpboost}{\ensuremath{\mathcal{C}}}
\newcommand{\alphaEinasto}{\ensuremath{\alpha}}
\newcommand{\Msol}{\ensuremath{M_{\odot}}}
\newcommand{\Msun}{\Msol}
\newcommand{\Mvir}{\ensuremath{M_{\mbox{{\footnotesize vir}}}}}
\newcommand{\Rvir}{\ensuremath{r_{\mbox{{\footnotesize vir}}}}}
\newcommand{\conc}{\ensuremath{c}}
\newcommand{\rhoc}{\ensuremath{\rho_{\mbox{{\footnotesize c}}}}}
\newcommand{\omegam}{\ensuremath{\Omega_{m}}}
\newcommand{\vcirc}{\ensuremath{v_{\mbox{{\footnotesize c}}}}}
\newcommand{\vmax}{\ensuremath{v_{\mbox{{\footnotesize max}}}}}
\newcommand{\rmax}{\ensuremath{r_{\mbox{{\footnotesize max}}}}}
\newcommand{\dSph}{dSph}
\newcommand{\dSphs}{dSphs}
\newcommand{\SDSS}{SDSS}
\newcommand{\degree}{^\circ}
\newcommand{\Jtilde}{\ensuremath{\tilde{J}}}

@@

In this paper we quantify the magnitude of this new substructure boost
and discuss its implications for the detection of dark matter
annihilation from subhalos in the Galactic halos. As we will show, the
total combined boost from Sommerfeld enhancement and substructure can
reach very large values (10$^5$ -- 10$^9$), and these boosts are only
weakly dependent on the assumptions that we have to make about the
abundance of substructure and the \DM\ density profile. We use these
total boosts together with published \DM\ density profiles of some of
the dwarf Spheroidal satellite galaxies of the Milky Way to predict
the total flux of gamma-rays from \DM\ annihilation that would be
detectable by the Fermi Gamma-ray Space Telescope (\Fermi) and find
that in most of the models that posit the Sommerfeld enhancement the
\DM\ annihilation signal from the dwarf Spheroidals shines above the
diffuse extragalactic gamma-ray background, which is the only
background signal at the high Galactic lattitudes of these satellite
galaxies. Finally, we also consider the prospects for detecting
smaller subhalos of the Galaxy by computing annihilation fluxes of
subhalos in the \VL\ II simulation. We will see that whether these
subhalos will be detectable in the near future is highly dependent on
the exact details of the Sommerfeld enhancement.

@@

\begingroup
%\squeezetable
\begin{table*}
\begin{ruledtabular}
\begin{tabular}{lrr@{.}lr@{}lr@{}lr@{.}lr@{.}lr@{.}l@{ $\times$ }lc}
dSph & \dist\ & \multicolumn{2}{c}{$\sigma_0$}\ & \multicolumn{2}{c}{\rmax} & \multicolumn{2}{c}{\vmax} & \multicolumn{2}{c}{\rhos} & \multicolumn{2}{c}{\rs} & \multicolumn{3}{c}{$M(<\rs)$} & References\\
& [kpc] & \multicolumn{2}{c}{[km s$^{-1}$]} & \multicolumn{2}{c}{[kpc]} & \multicolumn{2}{c}{[km s$^{-1}$]} & \multicolumn{2}{c}{[$\Msun$ kpc$^{-3}$]} & \multicolumn{2}{c}{[kpc]} & \multicolumn{3}{c}{[$\Msun$]}&\\[4pt]
\hline
Carina.......................................................................... & 101 & 6&8 & 3&\phantom{f}      & 16&$\!\!\!$      & 1&1 $\times$ $10^7$ & 1&4 & 7&4&$10^8$ & 1,3\\
Coma Berenices...........................................................      & 44  & 4&6 & $\cdot$&& $\cdot$& & 3&0 $\times$ $10^8$ & 0&3 & 2&0&10$^7$ & 2,4\\
Draco........................................................................... & 80  & 5&5 & 8&      & 40&      & 1&0 $\times$ $10^7$ & 3&7 & 1&2&$10^9$ & 1,3\\
Fornax.........................................................................  & 138 & 11&1& 4&      & 20&      & 1&0 $\times$ $10^7$ & 1&9 & 1&5&$10^8$ & 1,3\\
Leo I............................................................................  & 250 & 8&8 & 6&      & 30&      & 1&0 $\times$ $10^7$ & 2&8 & 5&2&$10^8$ & 1,3\\
Leo II...........................................................................  & 205 & 6&7 & 4&      & 20&      & 1&0 $\times$ $10^7$ & 1&9 & 1&5&$10^8$ & 1,3\\
Sculptor.......................................................................  & 79  & 8&5 & 6&      & 30&      & 1&0 $\times$ $10^7$ & 2&8 & 5&2&$10^8$ & 1,3\\
Sextans........................................................................  & 86  & 5&8 & 2&      & 10&      & 1&0 $\times$ $10^7$ & 0&9 & 1&9&$10^7$ & 1,3\\
Ursa Major II..............................................................   & 32  & 6&7 & $\cdot$&& $\cdot$& & 3&0 $\times$ $10^8$ & 0&3 & 2&0&$10^7$ & 2,4\\
Ursa Minor..................................................................    & 66  & 15&0& 6&      & 30&      & 1&0 $\times$ $10^8$ & 2&8 & 5&2&$10^8$ & 1,3\\
%Willman I............. & 38 &&& 5&5 $\times$ $10^8$ & 0&2 & 1&10$^{15}$\\
%Segue I.................. & 23 &&&&
\end{tabular}
\end{ruledtabular}
\caption{Properties of the dwarf Spheroidals used in this study.\\
References: (1) \citet{2008ApJ...672..904P}, (2) \citet{2008ApJ...678..614S}, (3) \citet{1998ARA&A..36..435M}, (4) \citet{2007ApJ...670..313S}.}\label{tab:dsphs}
\end{table*}
\endgroup

@@

At this point we can also make a connection to another quantity that
is often used to describe the astrophysical contribution to the \DM\
annihilation signal, $J$. This quantity is generally defined as
\begin{equation}
J \equiv \frac{1}{\Delta \Omega} \int_{\Delta \Omega} \dd \Omega\, \int_{\mbox{{\footnotesize los}}} \dd l\, \rho^2\, ,
\end{equation}
and is therefore equal to the line-of-sight integral over the density
squared averaged over solid angle. We can define a slight variant of
this which takes into account the Sommerfeld enhancement
\begin{equation}
\Jtilde \equiv \frac{1}{\Delta \Omega} \int_{\Delta \Omega} \dd \Omega\, \int_{\mbox{{\footnotesize los}}} \dd l\, \rho^2 \somm\, ,
\end{equation}
in which \somm\ is the Sommerfeld enhancement averaged over the
distribution of relative velocities. If we work this out for the
dwarf Spheroidals, using the expression from \eqnname\
(\ref{eq:lumrhosrsNFW}) for the line-of-sight integral averaged over
the solid angle spanned by the scale radius \rs\ we find that this
\Jtilde\ is approximately equal to the total boost \totalboost, with a
pre-factor \order(1). Therefore in the case of the \dSphs\ studied
here the \Jtilde\ can be approximately read off from the top panels of
Figure \ref{fig:subboost}.

It is clear from Figure \ref{fig:dsph} that the \DM\ annihilation flux
predicted to be coming from the \dSphs\ is very large in a large class
of models. We have to stress here that our predicted fluxes are all
for the particle physics specific parameters given in Section
\ref{sec:partphys}, \ie, \sigmaannv\ = $3 \times 10^{-26}$ cm$^3$
s$^{-1}$, \mdm\ = 700 GeV and a particular form for the \DM\
annihilation spectrum. To predict the gamma-ray flux for a particular
model with different values of these particle physics parameters one
is required to calculate the particle physics parameter specific
factor to the \DM\ annihilation flux and appropriately rescale the
values given here. In order to compare these signals to the expected
background flux we will only consider contributions to the background
flux of extragalactic gamma rays. Although there are also
contributions to the background flux from pion-decay, inverse Compton
scattering, and Bremsstrahlung, at the high Galactic lattitudes at
which the dwarf Spheroidals are found these are all of the same order
or smaller than the extragalactic background
\cite{2004ApJ...613..962S}. We have integrated the extragalactic gamma
ray flux as measured by EGRET over the energy range between 4 GeV --
250 GeV and multiplied it with the solid angle that the \dSphs\ span
on the sky. The extragalactic gamma ray flux is given by a power law
\cite{1998ApJ...494..523S}
\begin{equation}
\frac{\dd \Ngamma}{\dd E \,\dd A \, \dd t\,\dd \Omega} = k\,\left(\frac{E}{E_0}\right)^{-\alpha}\, ,
\end{equation}
where $k = 7.3 \times 10^{-6}$ cm$^{-2}$ s$^{-1}$ sr$^{-1}$
GeV$^{-1}$, $\alpha$ = 2.10, and $E_0 = 0.45$ GeV. We have included
the background flux of extragalactic photons in Figure
\ref{fig:dsph}. This background flux is in all cases of the order of
10$^{-9}$ -- 10$^{-10}$ photons cm$^{-2}$ s$^{-1}$, such that the \DM\
annihilation boosted by substructure and Sommerfeld enhancement
outshine this background in most of the parameter space, and by a few
orders of magnitude in most models. Given that the flux above 100 MeV
over a 55 days time-span needed to separate a point source of \DM\
annihilation from the background by \Fermi\ to give a 5$\sigma$
detection is about 10$^{-8}$ photons cm$^{-1}$ s$^{-1}$ and that the
necessary fluxes scale roughly as $t_{\mbox{{\footnotesize
exp}}}^{-1/2}$ \cite{2008JCAP...07..013B}, the \DM\ annihilation of
these \dSphs\ should be detectable by \Fermi\ in many of the models
parameterized by ($\mv/\mdm,\alpha$). The Ursa Major II \dSph\ is the
only \dSph\ for which the \DM\ annihilation dominates the background
in all the regions of parameter space. This shows that the newly
discovered, most-dark-matter-dominated satellite galaxies are possibly
the best targets for the indirect detection of \DM. We conclude that
the prospects for detecting \DM\ annihilation from the \dSphs\ with
\Fermi\ are very good in models that include a significant Sommerfeld
enhancement.

@@

{\bf Acknowledgements:} It is a pleasure to thank Ilias Cholis, Mulin
Ding, David Hogg, and Neal Weiner for comments and assistance. This
research was partially supported by NASA (grant NNX08AJ48G).
